Measuring contrast processing in the visual system using SSVEP

### 2024-05-13 Rversion/reviewR.tex
Added R version in subfolder

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for at least 100 years. One of the most
productive experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high SNR, and they allow us to analyze interactions between
stimulus components using a convenient mathematical framework. Here we
describe how SSVEPs have been used to study visual contrast over the
past 70 years (Dawson, 1954). Because our thinking about SSVEPs is
well-described by simple mathematical models, we embed code that
illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as contrast, sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Campbell \& Green, 1965),
defined as the inverse of the lowest contrast that can be reliably
detected. But neural responses can also be measured more directly using
techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG
method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (McKeefry \emph{et al.}, 1996;
Baseler \& Sutter, 1997; Di Russo \emph{et al.}, 2001\emph{a}).

Contrast is relatively simple to define: typically, it is specified as
the percentage deviation of a uniform stimulus from the background. So,
for example, a disk of 100 units of cone activation
(\(I_{\mathrm{stim}}\)) surrounded by a `background' of 50 units of
activation (\(I_{\mathrm{background}}\)) has a contrast of
\(\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}\)
= 100\%. Where patterns are more complex (for example, the sine-wave
gratings or Gabor patches common in vision science), the Michelson
(1927) definition of contrast is specified by the maximum and minimum
excursions from the mean:

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}).

@@

Because it provides a direct read-out of neural population activity, the
SSVEP signal can reveal key features of neural signal transduction. For
example, by varying the peak stimulus contrast parametrically, a
`contrast vs response' function can be measured - where the `response'
is typically defined as the amplitude of the SSVEP frequency component
at the stimulus frequency, or a low multiple thereof. This corresponds
closely to similar functions reported by studies measuring single unit
activity or local field potentials in the cortex (Shapley \& Victor,
1980; Morrone \emph{et al.}, 1982). However the SSVEP has the advantage
that it is non-invasive, and so can be measured in awake, behaving human
participants.

@@

Neurons have a limited dynamic range, yet they can transmit information
about visual stimuli that span many orders of magnitude. In the domain
of contrast, to some extent this is accomplished at a population level -
individual neurons typically implement a non-linear, sigmoidal contrast
vs response (CRF) transducer (Tolhurst \emph{et al.}, 1981; Albrecht \&
Hamilton, 1982) and different neurons exhibit peak sensitivity (defined
as the maximum slope of the function) at different contrast levels
(Carandini \& Heeger, 1994; Carandini \emph{et al.}, 1998; Busse
\emph{et al.}, 2009). A neuronal population will therefore span a
sensitivity range greater than any individual member.

@@

To better understand these processes, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we parametrically vary the
stimulus contrast (i.e.~the amplitude of the waveform), the amplitude of
the F1 component increases linearly with contrast (right panel).

@@

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual gain control using SSVEP data. For example, Tsai
\emph{et al.} (2012) demonstrated that a gain control model gave a good
account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. Our own work
on signal combination across eyes and space similarly demonstrated close
correspondence between the predictions of a computational model and
empirical data in humans (Baker \& Wade, 2017). More detailed modelling
of intracortical recordings (Groen \emph{et al.}, 2022) has revealed
details of the timecourse of gain control effects, specifically that
normalization is delayed slightly relative to the initial visual
response.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels, and extrapolaKting back along the
function (either contrast vs response measured at a constant spatial
frequency or spatial frequency vs response at a constant contrast level)
to estimate its intercept with the x-axis (see
Figure~\ref{fig-sweepvep}). This contrast level was shown to correspond
approximately with psychophysically measured detection thresholds
(Norcia \emph{et al.}, 1986).

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrasr levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Tyler \emph{et al.}, 1979). To avoid
hysteresis effects, the sweep is sometimes conducted both up and down in
the same experiment Norcia \& Tyler (1985\emph{a}). The sweep VEP
(really, a sweep `SSVEP') technique is now commonly-used to obtain a
rapid and objective measurement of visual acuity. In particular, because
of its relative speed and simplicity, this technique has now become a
standard for conducting tests of visual acuity in very young subjects or
where behavioural tests are not appropriate (Ridder, 2004; Bach \emph{et
al.}, 2008; Hoffmann \emph{et al.}, 2017; Bach \& Farmer, 2020).

@@

Recent work on understanding Parkinson's disease has also used
\emph{Drosophila} genetic models. Afsari \emph{et al.} (2014) found that
mutant flies produced stronger SSVEP responses to flickering lights than
control flies (see Figure~\ref{fig-plotclinical}d). The authors
theorised that differences in early gain control during development
might lead to visual deficits later in life. Although visual responses
are a convenient assay of neural function, it is likely that same
general process applies throughout the whole brain, including in the
motor system where the core Parkinson's symptoms (tremor, rigidity, slow
movement) manifest. The SSVEP differences were reduced by a kinase
inhibitor that targets the dopamine system, demonstrating how model
organisms can be used to test new pharmacological treatments. SSVEP
responses also provide a potential method to diagnose Parkinson's before
any symptoms manifest, and to monitor the effect of treatments.

@@

Attention exerts a profound influence on visual performance: for
example, allowing people to attend covertly to a spatial location
improves their performance on a target detection task significantly
(Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco \emph{et al.},
2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.}, 2004; Pestilli
\emph{et al.}, 2009). In principle, this enhancement may be driven both
by modulation of the underlying signal or noise characteristics, or by
additional decision-theoretic factors such as reduction in spatial
uncertainty (Petrov \emph{et al.}, 2006; Gould \emph{et al.}, 2007).
Early experiments with non-human primates showed little evidence for
attentionally-driven changes in neuronal spike rates (Luck \emph{et
al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et al.},
2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with the
advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (Morgan \emph{et al.}, 1996; Müller \emph{et al.}, 1998; Müller
\& Hillyard, 2000; Ding \emph{et al.}, 2006).

@@

One widespread recent application of the SSVEP technique is in the
design of brain-computer interfaces, which seek to control some aspect
of a computer using neural signals. The high SNR and precise frequency
resolution of SSVEPs make them an ideal candidate for this approach.
Typical studies may involve presenting an array of stimuli at different
contrast flicker frequencies, and having the participant select one
either by overt attention (i.e.~shifting fixation to foveate the
selected stimulus) or covert attention (i.e.~deploying attention to one
stimulus whilst keeping fixated) (Middendorf \emph{et al.}, 2000).
Because SSVEP signals are highly sensitive to both visual field position
(Di Russo \emph{et al.}, 2007; Ales \emph{et al.}, 2010) and attentional
state (Morgan \emph{et al.}, 1996; Müller \emph{et al.}, 1998; Lauritzen
\emph{et al.}, 2010; Verghese \emph{et al.}, 2012), the response to the
selected stimulus will typically increase relative to the others,
allowing it to be identified by an on-line algorithm. Because more than
one stimulus frequency is generally present, this modulation and the
associated changes in gain control will affect the entire pattern of
self- and intermodulation terms, allowing the choice to be decoded by a
multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to
distinguish from among a small set of possibilties: for example, in
early work, visual stimuli representing `Left' and `Right' commands in a
flight simulator were distinguished robustly (Middendorf \emph{et al.},
2000). Although SSVEP-based BCI interfaces typically do use contrast
flicker, work in this field has largely focused on optimising the
stimuli or decoders to increase decoding performance (e.g.~\ldots{} )
and reducing visual fatigue associated with the long-term presentation
of arrays of high-contrast flicker (Diez \emph{et al.}, 2024) rather
than studying visual contrast processing \textit{per se}. We therefore
note in passing that using SSVEPs to improve our understanding of early
contrast processing may yield benefits to this related field.

@@

The SSVEP is a powerful tool for studying contrast processing. It
provides a high SNR readout of neuronal activity that is unambigously
linked to the input. It is sensitive to both the amplitude and phase of
the input and when combined with source imaging, it can be extracted
from different cortical regions allowing researchers to track contrast
processing computations across the visual pathway. Because of the high
SNR, it can be measured in subjects where long recording durations are
impractical (for example, infants or patients with neurological
disorders) and at frequencies high enough to be effectively invisible
(Herrmann, 2001; Seijdel \emph{et al.}, 2023; Minarik \emph{et al.},
2023). The SSVEP is also able to `fingerprint' modulators of the inputs
through the harmonic and intermodulation terms they generate in the
output. In principle, each nonlinearity in the visual pathway can be
identified by its contribution to the frequency spectrum at different
recording locations (Regan \& Regan, 1988). This, it turn, allows
researchers to study how contrast processing depends on spatial and
temporal context and as well as changes in task, cognitive and
behavioural state and arousal.

@@

Finally, SSVEPs continue to provide a way of studying contrast in the
cortex. Here, we are often interested in how different visual parameters
interact. For example, how are signals from different eyes combined to
generate both scalar contrast values and also binocular depth cues? How
does contrast combination depend on the low level properties of the
individual inputs such as their retinotopic location, cone contrast,
spatiotemporal frequency, eye of origin and orientation? Colour vision
scientists are particularly interested in how chromatic signals
originating in a small number of cone-opponent retinal pathways are
transformed into a perceptual colour space where the `unique hues'
appear to be only weakly-related to the early retinal outputs and how
these computations are conditioned by the spatial and temporal
properties of the scene (Wandell, 1993; Gegenfurtner, 2003; Solomon \&
Lennie, 2007; Stoughton \& Conway, 2008; Kaneko \emph{et al.}, 2020; Li
\emph{et al.}, 2022). All of these questions can be addressed by using
SSVEP and frequency tagging to examine the computations that combine and
transform the inputs at different cortical stages (Busse \emph{et al.},
2009; Baker \& Wade, 2017; Katyal \emph{et al.}, 2018; Chen \&
Gegenfurtner, 2021; Watts \emph{et al.}, 2024). The original promise of
the SSVEP approach was that the entire complex-valued frequency spectrum
recorded at each lcoation provided detailed information about the
processing nonlinearities up to and including that point. In principle,
this would allow researchers to uniquely fit the parameters of their
computations models of visual processing (Regan \& Regan, 1988). To
date, the complexity of the neuronal computations at even the earliest
stage of visual processing have hampered this effort - researchers
typically restrict their analyses to the amplitudes of single, low-order
frequency components (for example, the inputs frequencies or simple sums
and differences of those frequencies). As our understanding of the early
visual system improves, it is becoming possible to generate more
realistic parameterised forward models of signal generation (Tsai
\emph{et al.}, 2012; Baker \& Wade, 2017; Chariker \emph{et al.}, 2020;
Schrimpf \emph{et al.}, 2020; Groen \emph{et al.}, 2022). Feeding
frequency-tagged inputs into these types of model allows us to generate
synthetic SSVEP responses that can be compared with those measured in
human subject. In principle, we are therefore able to use the SSVEP to
fit the parameters of early visual processing. This approach may also
allow us to develop more sensitive tests for the changes in early visual
processing that accompany a wide range of neurological diseases and
disorders.

>>>>

Contrast is the currency of the early visual system. Measuring the way
that the computations underlying contrast processing depend on factors
such as spatial and temporal frequency, age, clinical conditions,
eccentricity, chromaticity and the presence of other stimuli has been a
focus of vision science for over a century. One of the most productive
experimental approaches in this field has been the use of the
`steady-state visually-evoked potential' (SSVEP): a technique where
contrast modulating inputs are `frequency tagged' (presented at
well-defined frequencies and phases) and the electrical signals that
they generate in the brain are analyzed in the temporal frequency
domain. SSVEPs have several advantages over conventional measures of
visually-evoked responses: they have relatively unambiguous ouput
measures, a high SNR, and they allow us to analyze interactions between
stimulus components using a convenient mathematical framework. Here we
describe how SSVEPs have been used to study visual contrast over the
past 70 years (Dawson, 1954). Because our thinking about SSVEPs is
well-described by simple mathematical models, we embed code that
illustrates key steps in the modelling and analysis. This paper can
therefore be used both as a review of the use of SSVEP in measuring
human contrast processing, and as an interactive learning aid.

@@

Neurons in the visual areas of the brain are primarily responsive to
changes in cone photoreceptor activations across time and space. This
property, referred to as `contrast', sets the fundamental limits of our
visual abilities, which remain steady over a remarkably wide range of
environmental light levels. The human response to contrast can be
studied using many different techniques. Early work used psychophysical
methods to measure contrast sensitivity (Campbell \& Green, 1965),
defined as the inverse of the lowest contrast that can be reliably
detected. But neural responses can also be measured more directly using
techniques such as fMRI, MEG, and EEG. Here we will describe how an EEG
method known as the steady state visually evoked potential (SSVEP)
technique has contributed to our understanding of human contrast
processing in health, disease and throughout development.

@@

Contrast is one of the most fundamental pieces of information that the
eye transmits to the brain. It can be defined as the change in cone
photoreceptor activity over space (`spatial contrast') or time
(`temporal contrast'). Cone photoreceptors - which drive precortical
opponent pathways - contribute to both chromatic and achromatic
contrast, and although most of the research we describe here focuses on
achromatic contrast, SSVEPs have proven to be an excellent measure of
early chromatic processing as well (McKeefry \emph{et al.}, 1996; Di
Russo \emph{et al.}, 2001\emph{a}) (see also (Baseler \& Sutter, 1997)).

Contrast is typically specified as the percentage deviation of a uniform
stimulus from the background. So, for example, a disk of 100 units of
cone activation (\(I_{\mathrm{stim}}\)) surrounded by a `background' of
50 units of activation (\(I_{\mathrm{background}}\)) has a contrast of
\(\frac{I_{\mathrm{stim}} - I_{\mathrm{background}}}{I_{\mathrm{background}}}\)
= 100\%. Where patterns are more complex (for example, the sine-wave
gratings or Gabor patches common in vision science), the Michelson
(1927) definition of contrast is specified by the maximum and minimum
excursions from the mean:

@@

Although its mathematical definition is straightforward, the
computations that underlie contrast processing in the brain have been
the subject of intense research for many decades. The neural code for
contrast, even in the earliest parts of visual cortex, is not simply a
linear transform of the contrast at the retina - instead, contrast
signals undergo a cascade of nonlinear processing stages that, broadly,
attempt to normalise the output relative to the spatiotemporal
environment. This normalization, achieved through a computation called
`contrast gain control' (Heeger, 1992; Foley, 1994; Carandini \& Heeger,
2011) maximises the sensitivity of the visual system by making optimal
use of neuronal bandwidth. As an example, a grating placed at the centre
of a low-contrast background typically appears more intense than the
same grating when superimposed on a high contrast background (see
Figure~\ref{fig-centresurround}; note that the code used to produce all
figures in this review is available in python and R at:
https://github.com/wadelab/contrastReviewPaper).

@@

Because it provides a direct read-out of neural population activity, the
SSVEP signal can reveal key features of neural signal transduction. For
example, by varying the peak stimulus contrast parametrically, a
`contrast vs response' function (CRF) can be measured - where the
`response' is typically defined as the amplitude of the SSVEP frequency
component at the stimulus frequency, or a low multiple thereof. This
corresponds closely to similar functions reported by studies measuring
single unit activity or local field potentials in the cortex (Shapley \&
Victor, 1980; Morrone \emph{et al.}, 1982). However the SSVEP has the
advantage that it is non-invasive, and so can be measured in awake,
behaving human participants.

@@

Neurons have a limited dynamic range, yet they can transmit information
about visual stimuli that span many orders of magnitude. In the domain
of contrast, to some extent this is accomplished at a population level -
individual neurons typically implement a non-linear, sigmoidal CRF
transducer (Tolhurst \emph{et al.}, 1981; Albrecht \& Hamilton, 1982)
and different neurons exhibit peak sensitivity (defined as the maximum
slope of the function) at different contrast levels (Carandini \&
Heeger, 1994; Carandini \emph{et al.}, 1998; Busse \emph{et al.}, 2009).
A neuronal population will therefore span a sensitivity range greater
than any individual member.

@@

To understand these processes better, we will show how sinusoidal input
signals might be processed by the visual system to produce SSVEPs.
Figure~\ref{fig-linear} illustrates how sine waves of different
contrasts are processed in a linear system. The first panel shows the
input sine wave, which would be used to modulate stimulus amplitude over
time. Notice that there are five peaks in the waveform during the one
second sample, so the stimulation frequency is 5Hz (F1). The second
panel shows Fourier transform of the waveform, which contains a
substantial peak at this frequency. If we change the stimulus contrast
(i.e.~the amplitude of the waveform), the amplitude of the F1 component
increases linearly with contrast (right panel).

@@

The complexity of even a simple simulation of the frequency-domain
signal arising from non-linear interactions is intriguing. Presumably,
the signal measured from early visual cortex is the result of a cascade
of nonlinear retinal and cortical operations up to that point. It
therefore contains a `signature' or `fingerprint' of the computational
nature, order and parameters of those operations - including the shape
of the transducer functions and the computations involved in signal
combination. In principle, that information could be recovered from the
SSVEP signal - a possibility recognised in the early days of the
technique (Regan \& Regan, 1988). Although characterising the complete
set of computations along entire processing pathway is challenging,
careful parametric variation of the input stimuli does allow us to fit
models of early visual gain control using SSVEP data. For example, Tsai
\emph{et al.} (2012) demonstrated that a gain control model gave a good
account of the pattern of intermodulation responses produced by two
overlaid patterns flickering at different frequencies. This was achieved
by passing full stimulus waveforms through the transducer nonlinearity,
and calculating the Fourier spectrum of the model output. Our more
recent work on signal combination across eyes and space similarly
demonstrated close correspondence between the predictions of a
computational model and empirical data in humans (Baker \& Wade, 2017).
More detailed modelling of intracortical recordings (Groen \emph{et
al.}, 2022) has revealed details of the timecourse of gain control
effects, specifically that normalization is delayed slightly relative to
the initial visual response.

@@

Because SSVEP responses at detection threshold are very small,
estimating a threshold is achieved by measuring the contrast response
function at relatively high levels, and extrapolating back along the
function (either contrast vs response measured at a constant spatial
frequency or spatial frequency vs response at a constant contrast level)
to estimate its intercept with the x-axis (see
Figure~\ref{fig-sweepvep}). This contrast level was shown to correspond
approximately with psychophysically measured detection thresholds
(Norcia \emph{et al.}, 1986).

@@

A robust estimate of the threshold therefore requires the measurement of
the SSVEP amplitude at many different super-threshold contrast levels.
This was made faster by the development of the `sweep VEP' paradigm in
which the stimulus changed its contrast, spatial frequency or some other
property, throughout a trial (Tyler \emph{et al.}, 1979). To avoid
hysteresis effects, the sweep is sometimes conducted both up and down in
the same experiment (Tyler \emph{et al.}, 1979; Norcia \& Tyler,
1985\emph{a}; Norcia \emph{et al.}, 1990). The sweep VEP (really, a
sweep `SSVEP') technique is now commonly-used to obtain a rapid and
objective measurement of visual acuity. In particular, because of its
relative speed and simplicity, this technique has now become a standard
for conducting tests of visual acuity in very young subjects or where
behavioural tests are not appropriate (Ridder, 2004; Bach \emph{et al.},
2008; Hoffmann \emph{et al.}, 2017; Bach \& Farmer, 2020).

@@

Recent work on understanding Parkinson's disease has also used
\emph{Drosophila} genetic models. Afsari \emph{et al.} (2014) found that
mutant flies produced stronger SSVEP responses to flickering lights than
control flies (see Figure~\ref{fig-plotclinical}d). The authors
theorised that differences in early gain control during development
might lead to visual deficits later in life. Although visual responses
are a convenient assay of neural function, it is likely that the same
general process applies throughout the whole brain, including in the
motor system where the core Parkinson's symptoms (tremor, rigidity, slow
movement) manifest. The SSVEP differences were reduced by a kinase
inhibitor that targets the dopamine system, demonstrating how model
organisms can be used to test new pharmacological treatments. SSVEP
responses also provide a potential method to diagnose Parkinson's before
any symptoms manifest, and to monitor the effect of treatments.

@@

Attention exerts a profound influence on visual performance: for
example, instructing people to attend covertly to a spatial location
improves their performance on a target detection task significantly
(Bashinski \& Bacharach, 1980; Posner, 1980; Carrasco \emph{et al.},
2000; Cameron \emph{et al.}, 2002; Morrone \emph{et al.}, 2004; Pestilli
\emph{et al.}, 2009). In principle, this enhancement may be driven both
by modulation of the underlying signal or noise characteristics, or by
additional decision-theoretic factors such as reduction in spatial
uncertainty (Petrov \emph{et al.}, 2006; Gould \emph{et al.}, 2007).
Early experiments with non-human primates showed little evidence for
attentionally-driven changes in neuronal spike rates (Luck \emph{et
al.}, 1997; McAdams \& Maunsell, 1999; Mehta \emph{et al.},
2000\emph{a}, 2000\emph{b}; Marcus \& Van Essen, 2002), but with the
advent of spatially-resolved human brain imaging methods in the late
1990s it became apparent that spatial attention was linked to frank
changes in both fMRI BOLD responses (Tootell \emph{et al.}, 1998; Somers
\emph{et al.}, 1999; Gandhi \emph{et al.}, 1999; Brefczynski \& DeYoe,
1999; Kastner \emph{et al.}, 1999; Buracas \& Boynton, 2007; Silver
\emph{et al.}, 2007; Li \emph{et al.}, 2008; Murray, 2008) and EEG
signals (Morgan \emph{et al.}, 1996; Müller \emph{et al.}, 1998; Müller
\& Hillyard, 2000; Ding \emph{et al.}, 2006).

@@

One widespread recent application of the SSVEP technique is in the
design of brain-computer interfaces, which seek to control some aspect
of a computer using neural signals. The high SNR and precise frequency
resolution of SSVEPs are advantages for this approach. Typical studies
may involve presenting an array of stimuli at different flicker
frequencies, and having the participant select one either by overt
attention (i.e.~shifting fixation to foveate the selected stimulus) or
covert attention (i.e.~deploying attention to one stimulus whilst
keeping fixated) (Middendorf \emph{et al.}, 2000). Because SSVEP signals
are highly sensitive to both visual field position (Di Russo \emph{et
al.}, 2007; Ales \emph{et al.}, 2010) and attentional state (Morgan
\emph{et al.}, 1996; Müller \emph{et al.}, 1998; Lauritzen \emph{et
al.}, 2010; Verghese \emph{et al.}, 2012), the response to the selected
stimulus will typically increase relative to the others, allowing it to
be identified by an on-line algorithm. Because more than one stimulus
frequency is generally present, this modulation and the associated
changes in gain control will affect the entire pattern of self- and
intermodulation terms, allowing the choice to be decoded by a
multivariate pattern classifier.

This approach is primarily useful in situations that require the BCI to
distinguish from among a small set of possibilties: for example, in
early work, visual stimuli representing `Left' and `Right' commands in a
flight simulator were distinguished robustly (Middendorf \emph{et al.},
2000). Although SSVEP-based BCI interfaces typically do use contrast
flicker, work in this field has largely focused on optimising the
stimuli or decoders to increase decoding performance and reducing visual
fatigue associated with the long-term presentation of arrays of
high-contrast flicker (Diez \emph{et al.}, 2024), rather than studying
visual contrast processing \textit{per se}. We therefore note in passing
that using SSVEPs to improve our understanding of early contrast
processing may yield benefits to this related field.

@@

The SSVEP is a powerful tool for studying contrast processing. It
provides a high SNR readout of neuronal activity that is unambigously
linked to the input. It is sensitive to both the amplitude and phase of
the input and when combined with source imaging, it can be extracted
from different cortical regions allowing researchers to track contrast
processing computations across the visual pathway. Because of the high
SNR, it can be measured in subjects where long recording durations are
impractical (for example, infants or patients with neurological
disorders) and at frequencies high enough to be effectively invisible
(Herrmann, 2001; Seijdel \emph{et al.}, 2023; Minarik \emph{et al.},
2023). The SSVEP is also able to `fingerprint' modulators of the inputs
through the harmonic and intermodulation terms they generate in the
output. In principle, each nonlinearity in the visual pathway can be
identified by its contribution to the frequency spectrum at different
recording locations (Regan \& Regan, 1988). This, it turn, allows
researchers to study how contrast processing depends on spatial and
temporal context, as well as changes in task, cognitive and behavioural
state, and arousal.

@@

Finally, SSVEPs continue to provide a way of studying contrast in the
cortex. Here, we are often interested in how different visual parameters
interact. For example, how are signals from different eyes combined to
generate both scalar contrast values and also binocular depth cues? How
does contrast combination depend on the low level properties of the
individual inputs such as their retinotopic location, cone contrast,
spatiotemporal frequency, eye of origin and orientation? Colour vision
scientists are particularly interested in how chromatic signals
originating in a small number of cone-opponent retinal pathways are
transformed into a perceptual colour space where the `unique hues'
appear to be only weakly-related to the early retinal outputs and how
these computations are conditioned by the spatial and temporal
properties of the scene (Wandell, 1993; Gegenfurtner, 2003; Solomon \&
Lennie, 2007; Stoughton \& Conway, 2008; Kaneko \emph{et al.}, 2020; Li
\emph{et al.}, 2022). All of these questions can be addressed by using
SSVEP and frequency tagging to examine the computations that combine and
transform the inputs at different cortical stages (Busse \emph{et al.},
2009; Baker \& Wade, 2017; Katyal \emph{et al.}, 2018; Chen \&
Gegenfurtner, 2021; Watts \emph{et al.}, 2024). The original promise of
the SSVEP approach was that the entire complex-valued frequency spectrum
recorded at each location provided detailed information about the
processing nonlinearities up to and including that point. In principle,
this would allow researchers to uniquely fit the parameters of their
computational models of visual processing (Regan \& Regan, 1988). To
date, the complexity of the neuronal computations at even the earliest
stage of visual processing have hampered this effort - researchers
typically restrict their analyses to the amplitudes of single, low-order
frequency components (for example, the input frequencies or simple sums
and differences of those frequencies). As our understanding of the early
visual system improves, it is becoming possible to generate more
realistic parameterised forward models of signal generation (Tsai
\emph{et al.}, 2012; Baker \& Wade, 2017; Chariker \emph{et al.}, 2020;
Schrimpf \emph{et al.}, 2020; Groen \emph{et al.}, 2022). Feeding
frequency-tagged inputs into these types of model allows us to generate
synthetic SSVEP responses that can be compared with those measured in
human subjects. In principle, we are therefore able to use the SSVEP to
fit the parameters of early visual processing. This approach may also
allow us to develop more sensitive tests for the changes in early visual
processing that accompany a wide range of neurological diseases and
disorders.
